Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-22 doc/mc-paper.tex
Final edits for REB

We evaluate our toolchain using benchmark formulas from the 2022
standard and weighted model competitions.  Our tools handle all but
the largest dec-DNNF graphs generated by \dfour{}.  We also measure
the benefits of several optimizations as well as the relative
performance of the verified checker with one designed for high
performance and capacity.  We also show that our tools can provide
end-to-end verification of formulas that have been transformed by an
equivalence-preserving preprocessor.

@@

Capelli et al.\ developed a knowledge compiler
that generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{} program,
a modified version of \dfour{}, generates annotations to the
compiled representation, providing information about how the compiled version relates to the input clauses.
It also generates a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
Completing the certification involves running two different checkers on
the annotated dec-DNNF graph and the DRAT file.
Although the authors make informal arguments
regarding the soundness of their frameworks, these do not provide
strong levels of assurance.  Indeed, 
we have identified a weakness in the their
methodology due to an invalid assumption about the
guarantees provided by \dtrim{}, the program it uses to check the
DRAT file.

@@

This weakness would allow a buggy (or malicious) version
of \cdfour{} to generate a compiled representation of $\phi_2$ when
given $\phi_1$ as its input.  By adding the second clause to $R$,
the check with \dtrim{} would pass, as would the other tests performed
by their checker.  We have confirmed this possibility with
their compiler and checker\footnote{Downloaded May 18, 2023 as\\
\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}

This weakness can be corrected by restricting \dtrim{} to only add
clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
commandline argument to \dtrim{} that enforces this
restriction.\footnote{Available at
\url{https://github.com/marijnheule/drat-trim/tree/develop}.}  This weakness, however, illustrates the general challenge of
developing a new proof framework.
Without a formal verification, there are likely to be 
conditions that make the framework unsound.

@@

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.    Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In the supplement~\cite{bryant:sat:2023:supplement},
we also compare the performance of our toolchain to these two others.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated \emph{dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

@@

The \emph{function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
%%\begin{eqnarray}
\begin{displaymath}
\begin{array}{rcl}
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\ %\label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\ %\label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\ %\label{eqn:ring:sum}
\end{array}
\end{displaymath}
\end{proposition}
As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.

@@

@@

We can now define and compare the two related representations:
\begin{itemize}
\item A det-DNNF graph is a POG with negation applied only to the variables.
\item A dec-DNNF graph is det-DNNF graph with the further restriction that any
  sum node $\nodeu$ has have exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating dec-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

A CPOG file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$ implicitly adds an \emph{extension} variable $u$ and a set of \emph{defining} clauses $\theta_{u}$
  encoding the product or sum operation.
%  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
%  Generalizing to $k$ arguments would require
%  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then provides a way
of representing constant $0$.

@@

The defining clauses for a product or sum
operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
For the
extension variable $u$ associated with any POG node $\nodeu$, we can therefore
prove that any total assignment $\assign$ to the input variables which
also satisfies the POG defining clauses will
assign a value to $u$ such that $\assign(u) =
1$ if and only if $\assign \in \modelset(\nodeu)$.

@@

The supplement~\cite{bryant:sat:2023:supplement},
shows the CPOG description for
an input formula with five clauses, yielding a POG with five
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a dec-DNNF representation.
A dec-DNNF
graph can be directly translated  into a POG.
In doing this conversion,
our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so the only edges with negative polarity will have variables as children.
We can therefore view
view the POG as consisting
of \emph{literal} nodes corresponding to input variables and their negations, along with
\emph{nonterminal} nodes, which can be further classified as \emph{product} and \emph{sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
We can express the negation of $\phi_\noder$ in clausal form as
$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
when the formula $\inputformula \land \theta_{\obar{\noder}}$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a dec-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Most of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

\section{Implementations}
We have implemented programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}\footnote{The source code for all tools, along with a set of 
benchmark problems are available as part of the supplement\cite{bryant:sat:2023:supplement}}.  The proof generator is the same in both
cases, since it need not be trusted.
Our \emph{verified}
version of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our \emph{prototype} version is written in C\@.
It is faster and
more scalable, but we anticipate its need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

>>>>

We evaluate our toolchain using benchmark formulas from the 2022
standard and weighted model competitions.  Our tools handle all but
the largest dec-DNNF graphs generated by \dfour{}.  We also measure
the benefits of several optimizations as well as the relative
performance of the verified checker with one designed for high
performance and capacity.

We also show that our tools can provide end-to-end verification of
formulas that have been transformed by an equivalence-preserving
preprocessor.  That is, verification is based on the original formula,
and so proof checking certifies correct operation of the preprocessor,
the knowledge compiler, and the proof generator.

@@

Capelli, Lagniez, and Marquis developed a knowledge compiler
that generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{} program,
a modified version of \dfour{}, generates annotations to the
compiled representation, providing information about how the compiled version relates to the input clauses.
It also generates a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
Completing the certification involves running two different checkers on
the annotated dec-DNNF graph and the DRAT file.
Although the authors make informal arguments
regarding the soundness of their frameworks, these do not provide
strong levels of assurance.  Indeed, 
we have identified a weakness in the their
methodology due to an invalid assumption about the
guarantees provided by \dtrim{}, the program it uses to check the
DRAT file.

@@

This weakness would allow a buggy (or malicious) version
of \cdfour{} to generate a compiled representation of $\phi_2$ when
given $\phi_1$ as its input.  By adding the second clause to $R$,
the check with \dtrim{} would pass, as would the other tests performed
by their checker.  We have confirmed this possibility with
their compiler and checker.\footnote{Downloaded May 18, 2023 as\\
\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}

This weakness can be corrected by restricting \dtrim{} to only add
clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
commandline argument to \dtrim{} that enforces this
restriction.\footnote{Available at
\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
developing a new proof framework.
Without a formal verification, there are likely to be 
conditions that make the framework unsound.

@@

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.    Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In Section~\ref{sect:experimental} and the supplement~\cite{bryant:sat:2023:supplement},
we also compare the performance of our toolchain to these other two.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated \emph{dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight some special classes of Boolean formulas.  A formula is
  in \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

@@

The \emph{function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent,
and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{displaymath}
\begin{array}{rcl}
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\ 
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\ 
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\ 
\end{array}
\end{displaymath}
\end{proposition}
As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.

@@

We can now define and compare two related representations:
\begin{itemize}
\item A det-DNNF graph can be viewed a POG with negation applied only to variables.
\item A dec-DNNF graph is a det-DNNF graph with the further restriction that any
  sum node $\nodeu$ has exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating dec-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

@@

A CPOG file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$ implicitly adds an \emph{extension} variable $u$ and a set of \emph{defining} clauses $\theta_{u}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
%  Generalizing to $k$ arguments would require
%  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then provides a way
to represent constant $0$.

@@

The defining clauses for a product or sum
operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
For the
extension variable $u$ associated with any POG node $\nodeu$, we can therefore
prove that any total assignment $\assign$ to the input variables that
also satisfies the POG defining clauses will
assign a value to $u$ such that $\assign(u) =
1$ if and only if $\assign \in \modelset(\nodeu)$.

@@

The supplement~\cite{bryant:sat:2023:supplement},
shows the CPOG description for
an input formula with five clauses, yielding a POG with six
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a dec-DNNF representation.
A dec-DNNF
graph can be directly translated  into a POG.
In doing this conversion,
our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so the only edges with negative polarity will have variables as children.
We can therefore 
view the POG as consisting
of \emph{literal} nodes corresponding to input variables and their negations, along with
\emph{nonterminal} nodes, which can be further classified as \emph{product} and \emph{sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
We can express the negation of $\phi_\noder$ in clausal form as
$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
when the formula $\inputformula \land \theta_{\obar{\noder}}$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for general
use.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a dec-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$, requiring at most two RUP steps.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$ with a single RUP step,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Observe that 
all of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

\section{Implementations}
We have implemented programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}.\footnote{The source code for all tools
is available as part of the supplement\cite{bryant:sat:2023:supplement}}  The proof generator is the same in both
cases, since it need not be trusted.
Our \emph{verified}
version of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our \emph{prototype} version is written in C\@.
It is faster and
more scalable, but we anticipate its need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.
